Operational Feature Selection in Gaussian Mixture Models

### 2016-09-28 paper.tex
update

    The third type corresponds to the embedded method which do not separate the features selection process from the learning algorithm and allow interaction between the two processes. The basic example of such method is the decision tree algorithm in which a feature is selected for the creation of each node. Embedded methods also exist for other models, e.g. SVM \cite{guyon2002gene} \cite{weston2003use}.
²
    The selection method proposed in this work is a wrapper method associated to GMM models. Thus, in order to set a wrapper method, two elements are needed, a function to rank the various features which is defined in Section~\ref{sec:criterion} and a search strategy. Two search algorithms are presented: the sequential forward features selection method (Section \ref{sec:forward-presentation}) and the sequential floating forward feature selection method (\ref{sec:floating-presentation}) the second being a more complex variation of the first.

        \subsubsection{Criterion function}
        \label{sec:criterion}

@@

            \paragraph{Measures of correct classification}
            \label{sec:criterion-rate}

@@

            \paragraph{Similarity between distributions}

@@

        \subsubsection{Selection method}

            \paragraph{Sequential forward features selection}
            \label{sec:forward-presentation}

@@

            \paragraph{Sequential floating forward feature selection}
            \label{sec:floating-presentation}

@@

    The following classifiers are tested:
    \begin{itemize}
        \item A k-nearest-neighbors classifier (KNN) with OTB default parameters (32 as number of neighbors).
        \item A Random Forest classifier with parameters optimized by gridsearch (200 trees, 40 max depth, 50 size of subset of variables)%%TODO : pas clair ce qu'est "50 size of subset of variables"
        \item A GMM classifier with ridge regularization (GMM ridge) with regularization constant optimized by gridsearch.
    \end{itemize}
    The GMM classifier is part of the external module described in Section~\ref{sec:otb-module}.

@@

    %% TODO : c'est encore le cas ? à verifier: si on prends quand cela arréte de croitre rapidement il faut changer
    For SFS and  SFFS selection, the number of variables  to select is
    set to 30 for  the Aisa datset and 60 for  the Potsdam dataset and
    then  the  number  of  variables  which  maximizes  the  criterion
    function is used to perform classification.

@@

    %% TODO; Reprendre cette partie: ne parle pas à la fois de la
    %% classification et du temps de calcul. D'abors du commences à
    %% décrire et commenter sur la classification. Ensuite tu fais la
    %% même chose sur les temps de calculs. Et il faut en dire plus !!
    %%   Pour   les   classifs,   il  faut   parler   des   tests   de
    %%  significativité.  Pour les  temps  de  calculs, il  faut  plus
    %% détailler  pourquoi cela  va plus  ou moins  vite. Avec  FS, on
    %% réduit le nombre de variables et comme la régle de décision est
    %%  de compléxité  (d^3)  donc  quand on  réduit  d, on  améliorer
    %% grandement  les temps de  calcul, pour les RF,  plus il y  a de
    %% training samples, plus lse arbres vont êtres grands ...
    The  results show  that,  on this  dataset,  GMM classifiers  with
    feature selection  get the  best classification  rate.  Especially
    with  very  few training  samples,  GMM  classifiers with  feature
    selection outperform  the others.  In term  of computational time,
    the GMM classifiers  are as expected very  fast for classification
    and also  for training,  except when the  criterion function  is a
    classification rate.  Among the  three variations of the selection
    algorithm, none appears  to perform better than  the others. Using
    kappa  or Jeffries-Matusita  distance  as criterion  is equal  and
    using SFFS  does not give any  advantage. In this case,  using J-M
    distance  as criterion  and SFS  as  search strategy  is the  best
    choice in term of time efficiency.  % The computation of the kappa
                                       % seems  to suffer from  a lack
                                       %   of  parallelization   which
                                       %  could be  improved. ->  plus
                                       % tard dans la conclusion.

    Nevertheless,  good  performance  are also  obtained  with  Random
    Forest which becomes  more competitive when the  number of samples
    increases.  The KNN  classifier and the GMM  classifier with ridge
    regularization   are   totally   outperform  both   in   term   of
    classification rate and of processing time.

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 1000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu}}
        \begin{tabular}{lccccc}\toprule
          & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-6}
          GMM SFS kappa & 0.694 (0.002) & 0.669 (0.005) & 400 & 310 & 13.2 \\
          GMM SFS JM &    0.624 (0.028) & 0.631 (0.034) & 2 & 310   & 11 \\
          GMM SFFS JM &   0.624 (0.028) & 0.631 (0.034) & 2.6 & 310   & 11 \\
          GMM ridge &     0.632 (0.007) & 0.592 (0.010) & 10 & 2000 & all \\
          KNN &           0.637 (0.005) & 0.607 (0.005) &  &   & all \\
          Random Forest & 0.729 (0.004) & 0.673 (0.005) & 20 & 840  & all \\
          \bottomrule
        \end{tabular}
    \end{table}

    %% TODO: je te laisse modifier le tableau de la même manière (style, legende etc ...)
    \begin{table}[!t]
        \centering
        \caption{Results of classification with 50000 samples by class and 5 trials (standard deviation in parenthesis).\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{|l|c|c|c|c|c|}\hline
             & {\bfseries Kappa of} & {\bfseries Kappa of} & {\bfseries Training} & {\bfseries Classif.} & {\bfseries \# features} \\
             & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries time (s)} & {\bfseries time (s)} & {\bfseries to classify} \\ \hline

            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 460 & 2000 & all \\
            Random Forest & 0.851 (0.001) & 0.715 (0.001) & 2000 & 2000 & all \\
            \hline
        \end{tabular}
      \end{table}

      %% TODO: reprendre la discussion avec les commentaires de l'image précédente
    In this case, the classification rate of GMM classifiers with Jeffries-Matusita distance as criterion function drops clearly and has a high standard deviation. It could be due to the fact that the Gaussian hypothesis, on which this metric relies a lot, is not enough verified. When the classification rate is directly used as criterion, the classifier manages to select the relevant features. It has to be noticed that with kappa only 30 features are selected on average when with Jeffries-Matusita distance, the maximum number of features is selected almost always (set to 60).

    The KNN classifier and the GMM classifier with ridge regularization are again outperformed even if they get stable results.

    Finally, the Random Forest classifier and the GMM with kappa as criterion are from far the best classifiers. The only difference is in the processing time and also a small difference of classification rate with the training image. Random Forest is faster for the training process while the GMM classifier is faster for classifying.

>>>>

The third type corresponds to the embedded method which do not separate the features selection process from the learning algorithm and allow interaction between the two processes. The basic example of such method is the decision tree algorithm in which a feature is selected for the creation of each node. Embedded methods also exist for other models, e.g. SVM \cite{guyon2002gene} \cite{weston2003use}.

The selection method proposed in this work is a wrapper method associated to GMM models. Thus, in order to set a wrapper method, two elements are needed, a function to rank the various features which is defined in Section~\ref{sec:criterion} and a search strategy. Two search algorithms are presented: the sequential forward features selection method (Section \ref{sec:forward-presentation}) and the sequential floating forward feature selection method (\ref{sec:floating-presentation}) the second being a more complex variation of the first.

    \subsection{Criterion function}
    \label{sec:criterion}

@@

        \subsubsection{Measures of correct classification}
        \label{sec:criterion-rate}

@@

        \subsubsection{Similarity between distributions}

@@

    \subsection{Selection method}

        \subsubsection{Sequential forward features selection}
        \label{sec:forward-presentation}

@@

        \subsubsection{Sequential floating forward feature selection}
        \label{sec:floating-presentation}

@@

    The following classifiers are tested:
    \begin{itemize}
        \item A k-nearest-neighbors classifier (KNN) with OTB default parameters (32 as number of neighbors).
        \item A Random Forest classifier with parameters optimized by gridsearch (200 trees, 40 as max depth, 50 as size of the randomly selected subset of features at each tree node)
        \item A GMM classifier with ridge regularization (GMM ridge) with regularization constant optimized by gridsearch.
    \end{itemize}
    The GMM classifier is part of the external module described in Section~\ref{sec:otb-module}.

@@

    For SFS and  SFFS selection, the number of variables to select is set to 30 for the Aisa dataset and 60 for the Potsdam dataset. Then, a small processing is made to select the number of variables used for prediction. The evolution of the criterion function is retrieved from the model file and the discrete derivative of the criterion \emph{DJ} is computed and normalized by its maximum. The subset of variables used for classification $\Omega^{*}$ is the first subset for which this value is lower than a given threshold \emph{th}, equal  to 0.001\% in this experiment,
    \begin{equation*}
        \Omega^{*} = \Omega_{\min \{k|DJ_k<th\}}
    \end{equation*}
    with $DJ_k = \frac{J(\Omega_k) - J(\Omega_{k-1})}{\max_k (J(\Omega_k) - J(\Omega_{k-1}))}$.

@@

    The results show that, on this  dataset, GMM classifiers with feature selection get the best classification rate. Among the three variations of the selection algorithm, none appears to perform better than the others. Using kappa or Jeffries-Matusita distance as criterion is equal and using SFFS does not give any advantage.

    The difference with the second best classifier, Random Forest, appears to be significant when using 250 and 500 samples. Random Forest has similar performance in term of classification rate with 1000 samples. The GMM classifier with ridge regularization and the kNN classifier are both outperformed.

    In term of computational time, the GMM classifiers are as expected very fast for classification and also for training, except when the criterion function is a classification rate. In this case, using JM distance as criterion and SFS as search strategy is the best choice in term of time efficiency. The good performance in time can explain by the dimensional reduction. Actually, the decision rule corresponding to Equation~\ref{eq:decision} has a complexity in $d^3$ where $d$ is the dimension. Thus, reducing $d$ induces a reduction of the classification time.

    The processing times of the three standard classifiers suffer from the increase of training samples. For the GMM classifier with ridge, the selection of regularization parameter is more costly with more samples because the classification rate estimation needed. For the kNN classifier, the model stores all the training samples and the prediction implies to compute the distance to all the training samples which explains the increase of the processing time and additionally of the size of the model file. Finally, for the Random Forest classifier, the trees tends to be more deeper in order to capture the additional information available with more samples and that explains the increase of the processing time.
    % The computation of the kappa seems  to suffer from  a lack of  parallelization   which could be  improved. ->  plus tard dans la conclusion.

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 1,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu}}
        \begin{tabular}{lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-6}
            GMM SFS kappa & 0.694 (0.002)             & {\bfseries 0.669 (0.005)} & {\bfseries 0.533 (0.008)} & 400 & 310  & 13.2 \\
            GMM SFS JM &    0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2   & 310  & 11 \\
            GMM SFFS JM &   0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2.6 & 310  & 11 \\
            GMM ridge &     0.632 (0.007)             & 0.592 (0.010)             & 0.433 (0.008)             & 10  & 2000 & all \\
            KNN &           0.637 (0.005)             & 0.607 (0.005)             & 0.478 (0.002)             &     &      & all \\
            Random Forest & {\bfseries 0.729 (0.004)} & {\bfseries 0.673 (0.005)} & {\bfseries 0.529 (0.009)} & 20  & 840  & all \\
            \bottomrule
        \end{tabular}
    \end{table}

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 50,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{|lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-6}
            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 0.531 (0.005) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 0.440 (0.015) & 460 & 2000 & all \\
            Random Forest & {\bfseries 0.851 (0.001)} & {\bfseries 0.715 (0.001)} & {\bfseries 0.573 (0.002)} & 2000 & 2000 & all \\
            \bottomrule
        \end{tabular}
    \end{table}

    With this second dataset, the Random Forest classifier and the GMM classifier with kappa as selection criterion distinguished themselves by good performance. When using 1000 samples, no significant difference of classification rate has been observed on test set. But, with 50,000 samples, the Random Forest classifier better significantly better.

    It can be guessed that the fact that the Gaussian hypothesis is unlikely to be verified explains the difficulty to extract the finer information available in a larger dataset. The JM distance has stressed in Table~\ref{tab:crit} has a strong Gaussian hypothesis and the same hypotheses could also explain the drop of the classification rate of GMM classifiers with Jeffries-Matusita distance as criterion.

    The KNN classifier and the GMM classifier with ridge regularization are again outperformed even if they get more stable results than GMM classifiers with JM as criterion.

    The number of variables identified by the selection process shows that selection with JM distance as criterion identifies less relevant samples than with kappa as criterion. Moreover, the GMM classifier with kappa manages to get good performance with only 6.7\ of the variables with 1000 samples and 15\% with 50,000 samples.

    In term of processing time, results are similar than with the Aisa dataset. GMM classifiers with selection are very fast for prediction. For example, the GMM classifier with kappa as criterion is 63\% faster than the Random Forest classifier for prediction with 1000 samples and 83\% faster with 50,000 samples.
